---
title: The $\chi^2$ distribution
output: pdf_document
---
```{r, setup, include=FALSE}
source("../R/pdfs.R")
source("../R/cdfs.R")
source("../R/samplers.R")

library("functional")
library("ggplot2")
```

The [$\chi^2$ distribution](https://en.wikipedia.org/wiki/Chi-squared_distribution) is important in fields such as econometrics and life modelling. Parameterised by the degrees of freedom, $k$, the distribution function is $$F(x) = \frac{\gamma(\frac{k}{2}, \frac{x}{2})}{\Gamma(k/2)},$$ where $\gamma$ is the lower [incomplete gamma function](https://en.wikipedia.org/wiki/Incomplete_gamma_function), and probability density function $$f(x|k) = \frac{x^{k/2 - 1} e^{-x/2}}{2^{k/2}\Gamma(k/2)}.$$

The $\chi^2$ distribution is a special case of the gamma distribution with rate $\theta = \frac{1}{2}$ and shape $\frac{k}{2}$.

This distribution has mean $k$ and variance $2k$.

Since the sum of $k$ squared standard normal distributions are distributed as $\chi_k^2$, we can sample from the $\chi^2$ distribution by repeatedly sampling from a normal distribution and summing the squares of observations. Since the inverse of $F$ doesn't have a closed form, this is a computationally convenient method.

## Programs

We can sample from the gamma distribution with `chi_square_sampler`:

```{r}
k <- 1
n <- 200

chi_samples <- chi_square_sampler(n, k)
head(chi_samples)
```

R also has a built-in gamma distribution with pdf `dchisq`, which can be sampled with `rchisq`. We can test that this matches our implementation using a one-sample [Kolmogorov-Smirnov test](https://en.wikipedia.org/wiki/Kolmogorov%E2%80%93Smirnov_test):

```{r}
ks_results <- ks.test(chi_samples, "dchisq", df = k)
ks_results[["p.value"]]
```

In this instance, `r if (ks_results[["p.value"]]<0.05) {"since $p<0.05$, we conclude that the distributions match"} else {"either my code is broken or you experienced an exceptionally rare event"}`.

We can also plot the distributions against each other:
```{r}
dat <- data.frame(implementation =
                    factor(rep(c("chi_square_cdf", "dchisq"), each = n)),
                  value = c(chi_samples,
                            rchisq(n, df = k)))

ggplot(dat, aes(x = value, colour = implementation)) + geom_density()
```

Here, we have some noise from our relatively low number of samples, but the distributions clearly have the same shape.

## Problem 13

Write a program to generate a random sample of size $n$ from each of the following distributions:

- Chi-square with 1 degree of freedom ($\chi^2_1$);

- Chi-square with 5 degrees of freedom ($\chi^2_5$);

- Chi-square with 40 degrees of freedom ($\chi^2_{40}$).

Run your program for $n = 100, 300, 500$ and include a histogram in each case. How do these histograms change in shape as you change the degrees of freedom?

### Solution

For the 1 degree of freedom case, we get the following histograms:

```{r}
dat11 <- data.frame(cond = factor(rep("n=100", each = 100)),
                    value = chi_square_sampler(100, 1))
dat12 <- data.frame(cond = factor(rep("n=300", each = 300)),
                    value = chi_square_sampler(300, 1))
dat13 <- data.frame(cond = factor(rep("n=500", each = 500)),
                    value = chi_square_sampler(500, 1))

ggplot(rbind(dat11, dat12, dat13),
       aes(x = value, fill = cond)) +
  geom_histogram(binwidth = 0.5, alpha = 0.5, position = "dodge")
```

For the 5 degrees of freedom case, we get the following histogram:

```{r}
dat21 <- data.frame(cond = factor(rep("n=100", each = 100)),
                    value = chi_square_sampler(100, 5))
dat22 <- data.frame(cond = factor(rep("n=300", each = 300)),
                    value = chi_square_sampler(300, 5))
dat23 <- data.frame(cond = factor(rep("n=500", each = 500)),
                    value = chi_square_sampler(500, 5))

ggplot(rbind(dat21, dat22, dat23),
       aes(x = value, fill = cond)) +
  geom_histogram(binwidth = 1, alpha = 0.5, position = "dodge")
```

And for the 40 degrees of freedom case, we get the following histogram:

```{r}
dat31 <- data.frame(cond = factor(rep("n=100", each = 100)),
                    value = chi_square_sampler(100, 40))
dat32 <- data.frame(cond = factor(rep("n=300", each = 300)),
                    value = chi_square_sampler(300, 40))
dat33 <- data.frame(cond = factor(rep("n=500", each = 100)),
                    value = chi_square_sampler(500, 40))

ggplot(rbind(dat31, dat32, dat33),
       aes(x = value, fill = cond)) +
  geom_histogram(binwidth = 3, alpha = 0.5, position = "dodge")
```

Note that, as the degrees of freedom increase, the peak of the distribution shifts further right. Indeed, since the skew of the $\chi^2_k$ distribution is $\sqrt{\frac{8}{k}}$, we should expect the left tail to grow as $k$ increases. For high values of $k$, the skew will be approximately 0, so the distribution will have balanced left and right tails.